---
title: "Bootstrapping Diversity Indicies with Infer"
format: html
---

# Load packages

```{r}
#| label: packages

library(infer)
library(tidyverse)

theme_set(theme_bw(base_size = 16) + theme(aspect.ratio = 1)) # sets a global ggplot2 theme
```

# Generate fake data

```{r}
#| label: gen_data

set.seed(11)

species_levels <- paste0("sp_", 1:7)

fake_data <- tibble(
  environment = rep(c("env_a", "env_b"), each = 7),
    species = rep(species_levels, times = 2),
    n = c(
        18, 14, 12, 10, 9, 8, 0,   # env_a
        45, 9, 11, 13, 15, 17, 19   # env_b
    )
) |>
    tidyr::uncount(n)

fake_data
```

# Shannon diversity index

Shannon diversity index function for use with `infer::calculate()`. Copy this function into your document and run it so
that you can use with later.

```{r}
#| label: shannon_index

stat_shannon <- function(x, order, ...) {
  species <- x$species
  props <- table(species) / length(species)
  
  if (length(props) == 1) {
    return(0)
  }
  
  # Only sum over non-zero proportions to avoid log(0)
  h <- -sum(props[props > 0] * log(props[props > 0]))
  
  h
}
```

## Calculate observed

```{r}
#| label: calc_observed

obs_a <-
    fake_data |>
    filter(environment == "env_a") |>
    specify(response = species) |>
    calculate(stat = stat_shannon)

obs_b <-
    fake_data |>
    filter(environment == "env_b") |>
    specify(response = species) |>
    calculate(stat = stat_shannon)

```

## Boostrap sampling distribution

```{r}
#| label: bootstrap

boot_a <-
    fake_data |>
    filter(environment == "env_a") |>
    specify(response = species) |>
    generate(reps = 10000, type = "bootstrap") |>
    calculate(stat = stat_shannon)

boot_b <-
    fake_data |>
    filter(environment == "env_b") |>
    specify(response = species) |>
    generate(reps = 10000, type = "bootstrap") |>
    calculate(stat = stat_shannon)

boot_a |>
    visualise()

boot_b |>
    visualise()
```

## Confidence interval

```{r}
#| label: conf_int

ci_a <- 
    boot_a |>
    get_ci(level = 0.95, type = "percentile")

ci_b <- 
    boot_b |>
    get_ci(level = 0.95, type = "percentile")

ci_a
ci_b

boot_a |>
    visualise() +
    shade_ci(endpoints = ci_a)

boot_b |>
    visualise() +
    shade_ci(endpoints = ci_b)
```

## Plots

```{r}

obs_shannon <-
    bind_rows(
        env_a = obs_a,
        env_b = obs_b,
        .id = "environment"
    )

ci_shannon <- 
    bind_rows(
        env_a = ci_a,
        env_b = ci_b,
        .id = "environment"
    )

obs_shannon |>
    full_join(ci_shannon, by = join_by(environment)) |>
    ggplot(aes(x = environment, y = stat)) +
    geom_point(size = 4) +
    geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2)

```

# Simpson diversity index

Simpson diversity index function for use with `infer::calculate()`. Copy this function into your document and run it so
that you can use with later.

```{r}
#| label: simpson_index

stat_simpson <- function(x, order, ...) {
  species <- x$species
  props <- table(species) / length(species)
  
  # Simpson index: 1 - sum(p_i^2)
  d <- 1 - sum(props^2)
  
  d
}
```

## Calculate observed

```{r}
#| label: calc_observed_simpson

obs_simpson_a <-
    fake_data |>
    filter(environment == "env_a") |>
    specify(response = species) |>
    calculate(stat = stat_simpson)

obs_simpson_b <-
    fake_data |>
    filter(environment == "env_b") |>
    specify(response = species) |>
    calculate(stat = stat_simpson)

```

## Boostrap sampling distribution

```{r}
#| label: bootstrap_simpson

boot_simpson_a <-
    fake_data |>
    filter(environment == "env_a") |>
    specify(response = species) |>
    generate(reps = 10000, type = "bootstrap") |>
    calculate(stat = stat_simpson)

boot_simpson_b <-
    fake_data |>
    filter(environment == "env_b") |>
    specify(response = species) |>
    generate(reps = 10000, type = "bootstrap") |>
    calculate(stat = stat_simpson)

boot_simpson_a |>
    visualise()

boot_simpson_b |>
    visualise()
```

## Confidence interval

```{r}
#| label: conf_int_simpson

ci_simpson_a <- 
    boot_simpson_a |>
    get_ci(level = 0.95, type = "percentile")

ci_simpson_b <- 
    boot_simpson_b |>
    get_ci(level = 0.95, type = "percentile")

ci_simpson_a
ci_simpson_b

boot_simpson_a |>
    visualise() +
    shade_ci(endpoints = ci_simpson_a)

boot_simpson_b |>
    visualise() +
    shade_ci(endpoints = ci_simpson_b)
```

## Plots

```{r}

obs_simpson <-
    bind_rows(
        env_a = obs_simpson_a,
        env_b = obs_simpson_b,
        .id = "environment"
    )

ci_simpson <- 
    bind_rows(
        env_a = ci_simpson_a,
        env_b = ci_simpson_b,
        .id = "environment"
    )

obs_simpson |>
    full_join(ci_simpson, by = join_by(environment)) |>
    ggplot(aes(x = environment, y = stat)) +
    geom_point(size = 4) +
    geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2)

```
