---
title: "Uncertainty, sampling error and confidence intervals"
date: 2025-03-31
subtitle: "Lecture 3"
webr:
  packages:
    - tidyverse
    - infer
---

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}

```{webr}
#| edit: false
#| echo: false
#| output: false
options("readr.edition" = 1)

theme_set(theme_bw(base_size = 20) + theme(aspect.ratio = 1))

sample_trees <- read_csv("https://raw.githubusercontent.com/irmoodie/teaching_datasets/refs/heads/main/forest_data/forest_data_1.csv")

```

```{r}
#| echo: false
library(tidyverse)
library(infer)
library(patchwork)

theme_set(
  cowplot::theme_cowplot(font_family = "Atkinson Hyperlegible") +
    theme(aspect.ratio = 1, legend.position = "none")
)

tree_population <-
  read_csv(
    "/Users/iain/Documents/Projects/teaching_datasets/forest_data/forest_data.csv"
  ) |>
  mutate(
    x = runif(20000, 0, 1),
    y = runif(20000, 0, 1),
    tree_id = row_number()
  )

set.seed(123)
tree_samples <-
  tree_population |>
  rep_slice_sample(100, reps = 36)

forest_plot <- list()

forest_plot[["base"]] <-
  ggplot(tree_population, aes(x = x, y = y)) +
  theme_void() +
  theme(aspect.ratio = 1) +
  annotate(
    "rect",
    xmin = 0,
    xmax = 1,
    ymin = 0,
    ymax = 1,
    color = "black",
    fill = NA,
    linewidth = 0.5
  )

forest_plot[["pop"]] <-
  forest_plot[["base"]] +
  geom_point(size = 0.1, alpha = 0.5)

forest_plot[["pop_sample"]] <-
  forest_plot[["pop"]] +
  geom_point(
    data = tree_samples |> filter(replicate == 1),
    fill = "red",
    colour = "white",
    size = 2,
    shape = 23
  )

forest_plot[["all_samples"]] <-
  forest_plot[["base"]] +
  geom_point(
    data = tree_samples |> mutate(focal = if_else(replicate == 1, "a", "b")),
    aes(fill = focal),
    colour = "white",
    size = 2,
    shape = 23
  ) +
  scale_fill_manual(values = c("a" = "red", "b" = "gray50")) +
  facet_wrap(~replicate) +
  theme(strip.text = element_blank(), legend.position = "none")

##################################

sample_trees <- read_csv(
  "/Users/iain/Documents/Projects/teaching_datasets/forest_data/forest_data_1.csv"
)

sample_plots <- list()

sample_plots[["sample_dist"]] <-
  sample_trees |>
  ggplot(aes(x = tree_height_m)) +
  geom_histogram(bins = 12, colour = "white", fill = "grey70") +
  labs(
    title = "Sample distribution",
    x = "Tree height (m)",
    y = "Count"
  )

sample_plots[["sample_dist_obs"]] <-
  sample_plots[["sample_dist"]] +
  geom_vline(
    aes(xintercept = mean(tree_height_m)),
    colour = "red",
    linewidth = 1
  ) +
  geom_vline(
    aes(xintercept = mean(tree_height_m) + sd(tree_height_m)),
    colour = "blue",
    linewidth = 1
  ) +
  geom_vline(
    aes(xintercept = mean(tree_height_m) - sd(tree_height_m)),
    colour = "blue",
    linewidth = 1
  ) +
  annotate(
    "label",
    x = mean(sample_trees$tree_height_m),
    y = Inf,
    label = paste0("Mean = ", round(mean(sample_trees$tree_height_m), 2)),
    vjust = 1,
    colour = "red",
    size = 5
  ) +
  annotate(
    "label",
    x = mean(sample_trees$tree_height_m) + sd(sample_trees$tree_height_m),
    y = Inf,
    label = paste0("SD = ", round(sd(sample_trees$tree_height_m), 2)),
    vjust = 4,
    colour = "blue",
    size = 5
  ) +
  annotate(
    "segment",
    x = mean(sample_trees$tree_height_m),
    xend = mean(sample_trees$tree_height_m) + sd(sample_trees$tree_height_m),
    y = 0,
    yend = 0,
    colour = "blue",
    arrow = arrow(length = unit(0.2, "cm"), type = "closed", ends = "both")
  )
```

# Tree heights in a forest 

## Tree heights in a forest 
### Simulated population

::: {.columns}
::: {.column}
- Population $N$ = 20000
- Sample $n$ = 100
  - Trees sampled at random using a random number generator to provide coordinates
  - Closest tree to that coordinate was measured
:::
::: {.column}

```{r}
#| echo: false
forest_plot[["pop"]]
``
`
:::
:::

## Tree heights in a forest 
### Simulated population and sample(s)
`
``
{
  r
}
#| echo: false
#| fig-align: center
forest_plot[["pop_sample"]]
```

## Tree heights in a forest 
### Simulated population and sample(s)
```{r}
#| echo: false
#| fig-align: center
forest_plot[["all_samples"]]
```

# Worked example

## Tree heights in a forest 
### Sample distribution

```{r}
#| echo: false
#| fig-width: 10
forest_plot[["pop_sample"]] + sample_plots[["sample_dist"]]
```

## Tree heights in a forest 
### Observed statistics
::: {.columns}
::: {.column width=55%}

```{webr}
observed_mean <-
  sample_trees |>
  specify(response = tree_height_m) |>
  calculate(stat = "mean")

observed_sd <-
  sample_trees |>
  specify(response = tree_height_m) |>
  calculate(stat = "sd")
```
:::

::: {.column width=45%}
```{r}
#| echo: false
sample_plots[["sample_dist_obs"]]
```

:::
:::

## Tree heights in a forest 
### Uncertainty in our observed statistics

- If we took another random sample of 100 trees, it is unlikely that we would get exactly the same observed statistics
- We want to quantify this (**sampling error**)
- Problem: we (usually) only ever collect one sample
- Solution: generate more samples using information from our current sample

## Tree heights in a forest 
### Generating a bootstrap sample

- To generate more "samples", we use a re-sampling technique called the "bootstrap"
  - For original sample of size $n$, sample $n$ values **with replacement** many times
  - There are $n^n$ new samples we can generate from one sample
- Important: 
  - the original sample must be representative of the population
  - the original sample should be reasonably large (>>14, >30)

## Tree heights in a forest 
### Generating a bootstrap sample

```{webr}
#| warning: false
sample_trees |> # the original sample
  specify(response = tree_height_m) |>
  generate(reps = 1, type = "bootstrap") |> # generates 1 bootstrap sample
  ggplot(aes(x = tree_height_m)) +
  geom_histogram(bins = 16) + # plot the bootstrap sample as a histogram
  lims(x = c(0, 8)) # set limits so all plots on same scale
```

## Tree heights in a forest 
### Generating a bootstrap sample

```{r}
#| fig-width: 10
bootstrap_sample <-
  sample_trees |>
  specify(response = tree_height_m) |>
  generate(reps = 10000, type = "bootstrap")

sample_plots[["100_bootstrap"]] <-
  bootstrap_sample |>
  filter(replicate <= 100) |>
  ggplot(aes(x = tree_height_m)) +
  geom_histogram(bins = 12) +
  facet_wrap(~replicate) +
  theme_void() +
  theme(strip.text = element_blank(), legend.position = "none") +
  labs(title = "Bootstrap generated samples")

sample_plots[["sample_dist"]] + sample_plots[["100_bootstrap"]]

```

## Tree heights in a forest 
### Calculate statistics from bootstrap sample

```{r}
#| fig-width: 10
bootstrap_plots <- list()
bootstrap_plots[["mean_dist"]] <-
  bootstrap_sample |>
  calculate(stat = "mean") |>
  visualize() +
  labs(x = "Mean")

bootstrap_plots[["sd_dist"]] <-
  bootstrap_sample |>
  calculate(stat = "sd") |>
  visualize() +
  labs(x = "Standard Deviation")

bootstrap_plots[["mean_dist"]] + bootstrap_plots[["sd_dist"]]

```

## Tree heights in a forest 
### Calculate statistics from bootstrap sample

```{r}
#| fig-width: 10
bootstrap_plots[["mean_dist_obs"]] <-
  bootstrap_plots[["mean_dist"]] +
  geom_vline(
    aes(xintercept = mean(sample_trees$tree_height_m)),
    colour = "red",
    linewidth = 1
  )

bootstrap_plots[["sd_dist_obs"]] <-
  bootstrap_plots[["sd_dist"]] +
  geom_vline(
    aes(xintercept = sd(sample_trees$tree_height_m)),
    colour = "blue",
    linewidth = 1
  )

bootstrap_plots[["mean_dist_obs"]] + bootstrap_plots[["sd_dist_obs"]]

```

## Tree heights in a forest 
### Use bootstrap sampling distribution to quanitify sampling error

- Calculate the **standard error** (SE)
  - Standard deviation of the sampling distribution
- Calculate a **confidence interval** (CI)
  - *If we repeated our experiment many times and calculated a X% CI each time, the X% CI’s would include the “true” value X% of the time.*
  - **SE method:** Assume the sampling distribution is a normal distribution (bell-curve), and use a formula to find the values which contain the middle X% of the distribution (valid for means and some other statistics)
  - **Percentile method:** The middle X% of the sampling distribution (valid for all statistics and shapes of sampling distributions*)

## Tree heights in a forest 
### Calculate 95% CI from bootstrap sampling distribution

```{r}
#| fig-width: 10
bootstrap_plots[["mean_dist_obs_ci"]] <-
  bootstrap_plots[["mean_dist_obs"]] +
  shade_ci(get_ci(bootstrap_sample |> calculate(stat = "mean")))

bootstrap_plots[["sd_dist_obs_ci"]] <-
  bootstrap_plots[["sd_dist_obs"]] +
  shade_ci(get_ci(bootstrap_sample |> calculate(stat = "sd")))

bootstrap_plots[["mean_dist_obs_ci"]] + bootstrap_plots[["sd_dist_obs_ci"]]

```

## Tree heights in a forest 
### Calculate 95% CI from bootstrap sampling distribution

```{r}
#| eval: false
#| echo: true
bootstrap_sample <-
  sample_trees |>
  specify(response = tree_height_m) |>
  generate(reps = 10000, type = "bootstrap")
```

```{r}
#| echo: true
bootstrap_sample |>
  calculate(stat = "mean") |>
  get_confidence_interval(type = "percentile")

bootstrap_sample |>
  calculate(stat = "sd") |>
  get_confidence_interval(type = "percentile")
```

## Tree heights in a forest 
### Unusual scenario: we have multiple samples
```{r}
#| echo: false
#| fig-align: center
forest_plot[["all_samples"]]
```

## Tree heights in a forest 
### Unusual scenario: we know the populations true parameters

```{r}
#| echo: false
#| fig-align: center
forest_plot[["pop"]]
```

## Tree heights in a forest 
### How well did our approach work?

- From single sample (usual scenario):
  - Calculate observed statistics
  - Generate bootstrap samples to create a sampling distribution
  - Calculate 95% CI from bootstrap sampling distribution
- From multiple samples:
  - Calculate observed statistics
  - Make a sampling distribution from observed statistics in each sample
  - Use that sampling distribution to calculate 95% CI
- Compare with population actual parameters

## Tree heights in a forest 
### How well did our approach work?

```{r}
#| fig-width: 10
tree_estimates <- read_csv("data/03/tree_estimates.csv")

student_plots <- list()

student_plots[["observed_means"]] <-
  ggplot(
    tree_estimates,
    aes(x = observed_mean, y = fct_reorder(name, observed_mean))
  ) +
  geom_point() +
  labs(y = "Name", x = "Observed mean") +
  coord_cartesian(xlim = c(2.4, 3.5))

student_plots[["observed_sd"]] <-
  ggplot(
    tree_estimates,
    aes(x = observed_sd, y = fct_reorder(name, observed_sd))
  ) +
  geom_point() +
  labs(y = "Name", x = "Observed sd") +
  coord_cartesian(xlim = c(0.6, 1.6))

student_plots[["observed_means"]] + student_plots[["observed_sd"]]
```

## Tree heights in a forest 
### How well did our approach work?

```{r}
#| fig-width: 10
student_plots[["observed_means_ownCI"]] <-
  student_plots[["observed_means"]] +
  geom_errorbarh(aes(xmin = l_95_ci_mean, xmax = u_95_ci_mean))


student_plots[["observed_sd_ownCI"]] <-
  student_plots[["observed_sd"]] +
  geom_errorbarh(aes(xmin = l_95_ci_sd, xmax = u_95_ci_sd))

student_plots[["observed_means_ownCI"]] + student_plots[["observed_sd_ownCI"]]
```

## Tree heights in a forest 
### How well did our approach work?

CI: *If we repeated our experiment many times and calculated a 95% CI each time, the 95% CI’s would include the “true” value 95% of the time.*

## Tree heights in a forest 
### How well did our approach work?

```{r}
#| fig-width: 10

student_plots[["true_mean"]] <-
  student_plots[["observed_means_ownCI"]] +
  geom_vline(xintercept = mean(tree_population$tree_height_m), colour = "red")

student_plots[["true_sd"]] <-
  student_plots[["observed_sd_ownCI"]] +
  geom_vline(xintercept = sd(tree_population$tree_height_m), colour = "blue")

student_plots[["true_mean"]] + student_plots[["true_sd"]]
```

## Confidence intervals
### General workflow:

Get observed statistics:

- `specify()` response (and explanatory) variable(s)
- `calculate()` observed statistic

Get CI:

- `specify()` response (and explanatory) variable(s)
- `generate()` bootstrap samples
- `calculate()` observed statistic in each sample
- `get_confidence_interval()`

## Confidence intervals
### Examples: mean
```{r}
#| echo: true
iris_data <-
  iris |>
  filter(Species == "setosa")

iris_data |>
  specify(response = Petal.Width) |>
  calculate(stat = "mean")

iris_data |>
  specify(response = Petal.Width) |>
  generate(reps = 10000, type = "bootstrap") |>
  calculate(stat = "mean") |>
  get_confidence_interval(type = "percentile")
```

## Confidence intervals
### Examples: difference in means
```{r}
#| echo: true
iris_data <-
  iris |>
  filter(Species == "setosa" | Species == "versicolor")

iris_data |>
  specify(response = Petal.Width, explanatory = Species) |>
  calculate(stat = "diff in means", order = c("setosa", "versicolor"))

iris_data |>
  specify(response = Petal.Width, explanatory = Species) |>
  generate(reps = 10000, type = "bootstrap") |>
  calculate(stat = "diff in means", order = c("setosa", "versicolor")) |>
  get_confidence_interval(type = "percentile")
```

## Confidence intervals
### Examples: correlation
```{r}
#| echo: true
iris_data <-
  iris |>
  filter(Species == "setosa")

iris_data |>
  specify(response = Petal.Width, explanatory = Petal.Length) |>
  calculate(stat = "correlation")

iris_data |>
  specify(response = Petal.Width, explanatory = Petal.Length) |>
  generate(reps = 10000, type = "bootstrap") |>
  calculate(stat = "correlation") |>
  get_confidence_interval(type = "percentile")
```

