---
title: "Statistical inference with bootstrap"
subtitle: "Exercise 4"
date: 2026-03-27
execute:
  echo: true
  eval: false
---

# Get RStudio setup

Each time we start a new exercise, you should:

1. ~~Make a new folder in your course folder for the exercise (e.g. `biob11/exercise_4`).~~ Since we will continue
   working in the file from Exercise 3, you should keep working in the document initially.
2. Open RStudio
   - If you haven't closed RStudio since the last exercise, I recommend you do so and then re-open it. If it asks if you
     want to save your R Session data, choose no.
3. Set your working directory by going to *Session* -> *Set working directory* -> *Choose directory*, then navigate to
   the folder for this exercise.
4. ~~Create a new Rmarkdown document (*File* -> *New file* -> *R markdown..*). Give it a clear title.~~ Open the
   Rmarkdown document you made in exercise 3 (the `.Rmd` file, not the `.html` file)

You are now ready to start.

------------------------------------------------------------------------------------------------------------------------

# Peer review

To help you remember what you did in the last exercise, and to help you learn from each other, we will start today by
reviewing our code from Exercise 3.

::: {.callout-note icon="false"}
## ✅ Task

Sit in pairs or small groups (you need to be able to see each others screens). It might be a good idea to sit with
people who you didn't work with during the original exercise, but I won't police this.

Work your way through your documents together. Compare approaches and reasoning for data cleaning, which plots you made
and why, and how you answered the question at the end of the analysis.

If you find mistakes along the way, or you think one approach is better than another, help each other update their code
so it runs as desired.

:::

------------------------------------------------------------------------------------------------------------------------

# Tephritis phenotype II

## Getting setup

In the last exericse, we will used only the `tidyverse` package. In this exercise, we are going to use one additional
packaged called [`infer`](https://infer.tidymodels.org/){target="_blank"}.

::: {.callout-note icon="false"}
## ✅ Task

1. At the top of your document (but below the YAML frontmatter), make sure you load both the `tidyverse` and the `infer`
   package.
2. Run all the code cells in your document so that the data and its cleaned version have been loaded into R.

:::

## Average ovipositor

In this exercise, we are going to calculate a **95% confidence interval** around a sample statistic using a
**bootstrap** approach:

![The highlighted words in this diagram match with the functions you need to use to make your `infer` inference pipelines.](../slides/images/CI.svg)

Let's start with a simple question:

> What is the average ovipositor length of *Tephritis conura*?

::: {.callout-note icon="false"}
## ✅ Task

Make a new heading, called "Bootstrap analysis". Write answers to the following questions underneath:

1. What populations (statistical use of the word) are you trying to make inferences about? How broad do you think we can
   be?
2. How many samples^[Not observations] do we have?
3. What sample statistic should we use? Why is it approriate?
4. What information will a confidence interval provide us with?

:::

### Data

::: {.callout-note icon="false"}
## ✅ Task

In a new code cell, create a new dataset called `ovi_data` from your `clean_data` that only contains females, and has
all rows where `ovipositor_length_mm` was missing removed.

:::

### Plot

::: {.callout-note icon="false"}
## ✅ Task

1. In a new code cell, create a plot that shows the distribution of `ovipositor_length_mm` in the new `ovi_data`
   dataset. What sort of plot is more appropriate here?
2. Is this distribution a *sample* distribution or a *sampling*^[We should ban statisticians from naming things.]
   distribution? Write a short definition of each in your document.

:::

### Observed sample statistic

To calcualte our observed statistic, we are going to use the `infer` package. While you already have a way to calculate
statistics using `summarise()`, we will avoid this when performing statistical inferences. Use `summarise()` for making
pivot tables.

To use `infer` to calculate an observed statistic, we write:

```{r}
observed_ovi <- #<4>
  ovi_data |> #<1>
  specify(response = ______) |> #<2>
  calculate(stat = "______") #<3>

observed_ovi #<5>
```

1. Using the dataset we just made, we pipe `|>` it into the next line.
2. We `specify()` which columns we are interested in. In this case, since there's no explanatory variables, we just have
   a response.
3. We `calculate()` our chosen sample statistic. Try to use the [`infer`
   website](https://infer.tidymodels.org/){target="_blank"} or the helpfile for the `calculate()` function (write
   `?calculate` in your console) to figure out for yourself what you should write here.
4. We assign this to a new object called `observed_ovi`.
5. We write the name of the object so that its contents (a 1x1 dataframe) get printed into our document.

::: {.callout-note icon="false"}
## ✅ Task

In a new code cell, use the code above to calculate your desired sample statistic.

:::

### Bootstrap sampling distribution

Now we need a sampling distribution. To get that, we will use **bootstrap resampling**.

::: {.callout-note icon="false"}
## ✅ Task

Answer the following questions in your document before continuing:

1. Describe how we generate a single bootstrap sample?
2. What process are we mimicking by bootstrap resampling?
3. What strong, but often valid, assumption are we making when we use a bootstrap resampling process?

:::

To use `infer` to generate a bootstrap sampling distribution, we write:

```{r}
bootstrap_ovi <- #<2>
  ovi_data |> 
  specify(response = ______) |> 
  generate(reps = 10000, type = "bootstrap") |> #<1>
  calculate(stat = "______")
```

1. Notice how this is almost identical to the code we used to get the observed statistic, but we have one extra step,
   where we `generate()` 10000 bootstrap samples, then calculate the statistic for all of them.
2. We assign this to a new object, called `bootstrap_ovi`

::: {.callout-note icon="false"}
## ✅ Task

In a new code cell, use the code above to generate a bootstrap sampling distribution.

:::

### Confidence intervals

We can use the percentile method to calculate a 95% confidence interval.

::: {.callout-note icon="false"}
## ✅ Task

In your own words:

1. Describe how the percentile method derives a confidence interval.
2. What does the 95% refer to?
:::

`infer` has a helpful function to do this calculation for us:

```{r}
ovi_ci <-
  bootstrap_ovi |>
  get_confidence_interval(level = ______, type = ______) #<1>

ovi_ci 
```

1. Try to use helpfiles and the [`infer` website]((https://infer.tidymodels.org/){target="_blank"}) to figure out what
   you should write here.

::: {.callout-note icon="false"}
## ✅ Task

In a new code cell, use the code above to calculate a 95% confidence interval from your bootstrap sampling distribution
using the percentile method.

:::

### Visualising

`infer` has a helpful function to quickly plot a distribution of statistics. If you want more control on how it looks,
you can also use `ggplot()`

```{r}
bootstrap_ovi |>
  visualize()
```

::: {.callout-note icon="false"}
## ✅ Task

In a new code cell, use the code above to visualise your bootstrap simulated sampling distribution. Then answer the
following question:

1. What is on the X axis is this plot? What information does this plot show? Clarify why this plot shows different
   information to the one you made at the start of the exercise.

:::

We can shade the confidence interval we just made on this plot as well:

```{r}
bootstrap_ovi |>
  visualize() +
  shade_confidence_interval(ovi_ci) #<1>
```

1. Since the output of `visualize()` is a `ggplot`, we need to add layers using `+`.

::: {.callout-note icon="false"}
## ✅ Task

Visualise the confidence interval on top of the plot of the bootstrap simulated sampling distribution.
:::

### Answering the question

> What is the average ovipositor length of *Tephritis conura*?

::: {.callout-note icon="false"}
## ✅ Task

1. Write an answer to the question, referencing the results of your bootstrap analysis.
:::

------------------------------------------------------------------------------------------------------------------------

## Back to your own research question

Return to the question you answered at the end of Exercise 3 (where you had to choose 1 out of 6 to answer).

::: {.callout-note icon="false"}
## ✅ Final task

Calculate a **confidence interval** around a sample statistic using a **bootstrap** approach that helps to answer your
research question. Make notes explaining your thinking as you go.

Does your new analysis support or refute your previous conclusion?

Once you're done, again knit your document to a HTML file.

It will have been saved next to wherever your `.Rmd` file is saved.

Upload your `.html` file as your [assignment for this exercise in
Canvas](https://canvas.education.lu.se/courses/39432/assignments/290661).

:::
